Contracts for unions and intersections are hard for Nickel

### 2021-05-18 paper.tex
Rework the optimizations section and its examples

\lstset{
  language={nickel},
  basicstyle=\small\ttfamily, % Global Code Style
  captionpos=b, % Position of the Caption (t for top, b for bottom)
  extendedchars=true, % Allows 256 instead of 128 ASCII characters
  tabsize=2, % number of spaces indented when discovering a tab
  columns=fixed, % make all characters equal width
  keepspaces=true, % does not ignore spaces to fit width, convert tabs to spaces
  showstringspaces=false, % lets spaces in strings appear as real spaces
  breaklines=true, % wrap lines if they don't fit
  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
  frameround=tttt, % make the frame round at all four corners
  framesep=4pt, % quarter circle size of the round corners
  numbers=left, % show line numbers at the left
  numberstyle=\tiny\ttfamily, % style of the line numbers
  % commentstyle=\color{dark-grey}, % style of comments
  % keywordstyle=[1]\color{blue-portage}, % style of keywords
  % keywordstyle=[2]\color{orange-vivid-tangerine}, % style of keywords
  % keywordstyle=[3]\color{blue-portage}, % style of keywords
  % keywordstyle=[4]\color{green-pea}, % style of keywords
  % keywordstyle=[5]\color{pink-froly}, % style of keywords
  % stringstyle=\color{blue-marguerite}, % style of strings
}

@@

The performance of modern programs heavily relies on non-trivial optimizations
performed by the compiler or the interpreter. Even more so for functional
languages, whose operational semantics is usually very different from the actual
primitives of modern processors: naive code generation can exhibit unacceptable
slowdowns. Functional programs tend to make heavy use of functions, and a function
call is a costly operation from the point of view of low-level execution.

One such crucial optimization is inlining. Inlining eliminates a function
application by directly substituting the function for its definition at compile
time (or before execution, for interpreted language). This is especially
important for small functions that are called repeatedly.

\begin{lstlisting}[language=nickel]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}

Another optimization is common subexpression elimination (CSE). CSE somehow
goes into the other direction: it factors out equal subexpressions to avoid
recomputing them several times.

\begin{lstlisting}[language=nickel]
let elemAtOrLast = fun index list =>
  if index == lists.length list then
    lists.elemAt (lists.length list - 1) list
  else
    lists.elemAt index list
\end{lstlisting}

Similar transformations exist beyond CSE, such as loop-invariant code motion or
some of Haskell's let-floating~\ref{}, which amount to hoist invariant
expressions out from repeated blocks in order to avoid recomputing them. In
particular, take the following example:

\begin{lstlisting}[language=nickel]
let f = fun x =>
  g y (x + y)

// We want
let f =
  let _g = g y in
  fun x => \_g (x + y)
\end{lstlisting}

The soundness of these optimizations is tied the invariance of the semantics of
programs with respect to substitution. Inlining requires that one can replace
the application of a function by its body: this is $\beta$-reduction, and it is
hard to think of a language when it is not a valid transformations. On the other
hand, CSE-like transformations require an expansion property, of the form

$$M[N/x] \simeq let~x~=~M~in~M[x]$$

This does not necessarily holds, and is in particular trivially violated in
presence of side effects:

$$let~f~x~=~print~"hi";~(x+1)~in~(f~1, f~1) \\
\not \simeq let~f~x~=~print~"hi";~(x+1)~in~let~y~=~f~1~in~(y,y)$$

However, this property does hold for \emph{pure} terms, that are terms without
side-effects. Pure languages, or languages with a form of effect tracking such
as Haskell, are able to identify such pure terms. Other languages rely on
sub-approximations (for instance, applying optimizations to arithmetic
expressions).
\newpage

>>>>

\lstset{
  language={nickel},
  basicstyle=\small\ttfamily, % Global Code Style
  captionpos=b, % Position of the Caption (t for top, b for bottom)
  extendedchars=true, % Allows 256 instead of 128 ASCII characters
  tabsize=2, % number of spaces indented when discovering a tab
  columns=fixed, % make all characters equal width
  keepspaces=true, % does not ignore spaces to fit width, convert tabs to spaces
  showstringspaces=false, % lets spaces in strings appear as real spaces
  breaklines=true, % wrap lines if they don't fit
  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
  frameround=tttt, % make the frame round at all four corners
  framesep=4pt, % quarter circle size of the round corners
  numbers=left, % show line numbers at the left
  numberstyle=\tiny\ttfamily, % style of the line numbers
  escapeinside={(*@}{@*)} % escape sequence to insert latex
  % commentstyle=\color{dark-grey}, % style of comments
  % keywordstyle=[1]\color{blue-portage}, % style of keywords
  % keywordstyle=[2]\color{orange-vivid-tangerine}, % style of keywords
  % keywordstyle=[3]\color{blue-portage}, % style of keywords
  % keywordstyle=[4]\color{green-pea}, % style of keywords
  % keywordstyle=[5]\color{pink-froly}, % style of keywords
  % stringstyle=\color{blue-marguerite}, % style of strings
}

@@

The performance of modern programs heavily relies on non-trivial optimizations
performed by the compiler or the interpreter. Even more so for functional
languages, whose operational semantics is usually far removed from the actual
primitives of modern processors: naive execution can exhibit unacceptable
slowdowns. Functional programs tend to make heavy use of functions, and a
function call is a costly operation from the point of view of low-level
execution.

One crucial optimization is inlining(Figure~\ref{fig:optimizations-inlining-ex}). Inlining eliminates a function application
by directly substituting the function for its definition at compile time (or
before execution, for interpreted language). This is especially important for
small functions that are called repeatedly.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let subList = fun l1 l2 =>
  lists.any (fun x => x == elt)
  && subList (list.tail l1) l2
\end{lstlisting}
  \end{center}
\caption{Inlining}
\label{fig:optimizations-inlining-ex}
\end{figure}

Another optimization is common subexpression elimination (CSE,
Figure~\ref{fig:optimizations-cse-ex}). CSE goes into the other direction: it
factors out equal subexpressions to avoid recomputing them multiple times.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elemAtOrLast = fun index list =>
  if index > lists.length list - 1 then
    lists.elemAt (lists.length list - 1) list
  else
    lists.elemAt index list
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let elemAtOrLast = fun index list =>
  let l = lists.length list - 1 in
  if index > l then
    lists.elemAt l list
  else
    lists.elemAt index list
\end{lstlisting}
  \end{center}
\caption{Common subexpression elimination}
\label{fig:optimizations-cse-ex}
\end{figure}

Similar transformations exist beyond CSE, such as loop-invariant code motion or
some of Haskell's let-floating~\ref{}, which amount to hoist invariant
expressions out from repeated blocks in order to avoid recomputing them. In
particular, take the example of Figure~\ref{fig:optimizations-let-floating-ex}.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let f = fun x => g y (x + 1) 
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let f' = let g' = g y in
  fun x => g' (x + 1)
\end{lstlisting}
  \end{center}
\caption{Let-floating}
\label{fig:optimizations-let-floating-ex}
\end{figure}

The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
called. This may be costly, in particular in the presence of contracts: if
\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
on it, the penalty could be linear in the size of \lstinline+y+. A sensible
thing to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
Figure~\ref{fig:optimizations-let-floating-ex}, which is something Haskell's
let-floating transformations could indeed do, given \lstinline+g+ is pure.

The soundness of these optimizations is tied the invariance of the semantics of
programs with respect to some substitutive transformations\todo{bad naming}.
Inlining requires that one can replace the application of a function by its
body: this is $\beta$-reduction, and it is hard to think of a language where it
is not a valid transformations. On the other hand, CSE-like transformations
require an expansion property of the form

$$M[N/x] \simeq let~x~=~N~in~M[x]$$

This equation clearly fails in a call-by-value language with side effects:

\begin{lstlisting}[language=Nickel]
let f x = print "hi";(x+1) in (f 1,f 1) 
(*@ $\not \simeq$ @*) let f x = print "hi";(x+1) in
(*@ \phantom{$\not \simeq$} @*) let y = f 1 in (y,y)
\end{lstlisting}

However, this property does hold for \emph{pure} terms, that are terms without
side-effects. In a pure language, or a language with effect tracking like
Haskell, pure terms can be safely identified for applying optimizations. Other
languages rely on under-approximations.

\unsure{Should we tease already here that unions and intersections break this
even for pure terms?}
\newpage
